Dark Control: The Default Mode Network as a Reinforcement Learning Agent

### 2018-08-11 paper/paper.tex
some edits to summary

\subsection{Summary and empirical hypotheses}

@@

The current account of DMN has multiple empirical consequences and predictions. Perhaps one of the first experimental venue concerns the neural correlates of the Bellman equation in the DMN. There are already relationship between the decomposition of consecutive action choices by the Bellman equation and neuroscientific insights:
specific neural activity in the dorsal prefrontal cortex (BA9)
was for instance linked to processing ``goal-tree sequences''
in human brain-imaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further brain-imaging experiments corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while
maintaining information in working memory''
\citep{braver2002role}
and to functionally couple with the hippocampus conditioned by ``deep versus shallow planning''
\citep{kaplan2017neural}.
%
Moreover,
neurological patients with lesions in this DMN region were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
Here, we hypothesize a functional dissociation between computations pertaining to action policy versus adapting stimulus-value associations as we expect implementation in different subsystems of the DMN. First, we expect that fMRI signals in the right temporo-parietal junction relate to behavioral changes subsequent to adaptation in the action choice tendencies (policy matrix) involved in non-value-related prediction error. Second, fMRI signals in the ventromedial prefrontal cortex should relate to behavioral changes following adaptation in value estimation (value matrix) due to reward-related stimulus-value association. We finally expect that fMRI signals in the posteromedial cortex, as a potential global information integrator, are related to shifts in overt behavior based on previous adaptations in both policy or value estimation.

Beyond the Bellman equation and its potential underlying neural mechanisms, our process model of the DMN has also implication for experiments in neuroeconomy, especially for temporal discounting and continuous learning paradigms. Regarding the later, we for instance hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, the more the fMRI signals in the reward circuits should be independent of the reward context in the current sensory environment. In the case of temporal discounting, we hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits
\citep{luksys2009stress, haushofer2014psychology}.
Using a temporal discounting experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age and personality traits.
% To exam-ine this question, Fellows and Farah (2005) investigated the effects of dorsolateral frontal (DLF) and ventromedial frontal (VMF) lobe damage on two aspects of temporal foresight—namely, temporal discounting and the futuretime perspective (measuredby an adaptation of Wallace’s [1956] list of future life events).

@@

Another prediction from the current model is that the HC may contribute to generating
perturbed action-transition-state-reward samples
as batches of pseudo-experience
(i.e., recalled, hypothesized, and forecasted scenarios).
The small variations in these experience samplings allow searching
a larger space of model parameters and possible experiences.
Taken to its extreme, stochastic recombination of experience
building blocks can further optimize the behavior of the RL agent
by learning from scenarios in the environment that the agent might
encounter only very rarely or never.
An explanation is thus offered for experiencing seemingly familiar situations that
a human has however never actually encountered (i.e., d\'{e}j\`{a} vu effect).
While such a situation may not have been experienced in the physical world,
the DMN may have previously stochastically generated, evaluated, and adapted to
such a randomly synthesized event.
Generated representations arguably are ``internally manipulable,
and can be used for attempting actions internally,
before or instead of acting in the external reality,
and in diverse goal and sensory contexts, i.e. even outside the context in which they were learned''
\citep{pezzulo2011grounding}.
In the context of scarce environmental input and feedback
(e.g., mind-wandering or sleep),
mental scene construction allows pseudo-experiencing possible
future scenarios and action outcomes.

This interplay between memory and “mind-searching” also supports that experience replay for browsing problem solutions subserved by the DMN is contributing to choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before making analogous choices in the future. We hypothesize that, in addition to the hippocampus, there is a necessity of cortical DMN regions for “mind-searching” candidate actions during choice behavior. It can be experimentally corroborated by causal disruption of DMN regions, such as by circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.
From the perspective of a RL agent,
prediction in the DMN reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

A last empirical perspective concerns the role of synaptic epigenesis in the creation of new entry in policy matrix. Indeed, we did not adress here the additional layer of learning which concerns the addition of new entries in the state and action spaces. Extension of the action repertoire could be biologically realized by synaptic epigenesis \citep{gisiger_acquisition_2005}. Indeed, the tuning of synaptic weights through learning can stabilize additional patterns of activity by creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors can then constrain both the number of factors taken into account by decision processes and the possible behaviors of the agent \citep{wang_decision_2008}. To examine this potential higher-level mechanism, we propose to probe how synaptic epigenesis is related to neural correlates underlying policy matrix updates: in humans the changes of functional connectivity between DMN regions can be investigated following a temporal discounting experiment and in monkeys or rodents anterograde tracing can be used to study how homolog regions of the DMN present increased synaptic changes compare to other parts of the brain.

>>>>

\subsection{Summary and hypotheses for future studies}

@@

Our formal account on the DMN readily motivates several empirical predictions for future neuroscience research. Perhaps one of the first experimental venue concerns the neural correlates of the Bellman equation in the DMN. There are already relationship between the decomposition of consecutive action choices by the Bellman equation and neuroscientific insights:
specific neural activity in the dorsal prefrontal cortex (BA9)
was for instance linked to processing ``goal-tree sequences''
in human brain-imaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further brain-imaging experiments corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while
maintaining information in working memory''
\citep{braver2002role}
and to functionally couple with the hippocampus conditioned by ``deep versus shallow planning''
\citep{kaplan2017neural}.
%
Moreover,
neurological patients with lesions in this DMN region were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.

We therefore hypothesize a functional dissociation between computations pertaining to action policy versus adapting stimulus-value associations as we expect implementation in different subsystems of the DMN. First, we expect that fMRI signals in the right temporo-parietal junction relate to behavioral changes subsequent to adaptation in the action choice tendencies (policy matrix) involved in non-value-related prediction error. Second, fMRI signals in the ventromedial prefrontal cortex should relate to behavioral changes following adaptation in value estimation (value matrix) due to reward-related stimulus-value association. We further expect that fMRI signals in the posteromedial cortex, as a potential global information integrator, are related to shifts in overt behavior based on previous adaptations in both policy or value estimation.

Our process model of the DMN has also implications for experiments in neuroeconomy; especially for temporal discounting and continuous learning paradigms. More specifically, we hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a multi-step learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, the more the fMRI signals in the reward circuits should be independent of the reward context in the current sensory environment. In the case of temporal discounting, we hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits
\citep{luksys2009stress, haushofer2014psychology}.
Using such a delayed-reward experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age, education, IQ, and personality traits.
% To exam-ine this question, Fellows and Farah (2005) investigated the effects of dorsolateral frontal (DLF) and ventromedial frontal (VMF) lobe damage on two aspects of temporal foresight—namely, temporal discounting and the futuretime perspective (measuredby an adaptation of Wallace’s [1956] list of future life events).

@@

As another experimental prediction derived from our MDP approach to the DMN,
the HC may contribute to generating
perturbed action-transition-state-reward samples
as batches of pseudo-experience
(i.e., recalled, hypothesized, and forecasted scenarios).
The small variations in these experience samplings allow searching through
a larger space of model parameters and candidate experiences.
Taken to its extreme, stochastic recombination of experience
building blocks can further optimize the behavior of the RL agent
by learning from scenarios in the environment that the agent might
encounter only very rarely or never.
An explanation is thus offered for experiencing seemingly familiar situations that
a human has however never actually encountered (i.e., d\'{e}j\`{a} vu effect).
While such a situation may not have been experienced in the physical world,
the DMN may have previously stochastically generated, evaluated, and adapted to
such a randomly synthesized event.
Generated representations arguably are ``internally manipulable,
and can be used for attempting actions internally,
before or instead of acting in the external reality,
and in diverse goal and sensory contexts, i.e. even outside the context in which they were learned''
\citep{pezzulo2011grounding}.
In the context of scarce environmental input and feedback
(e.g., mind-wandering or sleep),
mental scene construction allows pseudo-experiencing possible
future scenarios and action outcomes.

A possible interplay between memory retrieval and ``mind-searching'' moreover suggests that experience replay for browsing problem solutions subserved by the DMN contributes to choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before making analogous choices in the future. We hypothesize that, in addition to the hippocampus, there is a necessity of cortical DMN regions for ``mind-searching'' candidate actions during choice behavior. It can be experimentally corroborated by causal disruption of DMN regions, such as by circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.
From the perspective of a RL agent,
prediction in the DMN reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

A last experimental prediction for future studies concerns the role of synaptic epigenesis in the creation of new entries in policy matrix. Indeed, we did not adress here the additional layer of learning which concerns the addition of new entries in the state and action spaces. Extension of the action repertoire could be biologically realized by synaptic epigenesis \citep{gisiger_acquisition_2005}. The tuning of synaptic weights through learning can stabilize additional patterns of activity by creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors can then constrain both the number of factors taken into account by decision processes and the possible behaviors of the agent \citep{wang_decision_2008}. To examine this potential higher-level mechanism, we propose to probe how synaptic epigenesis is related to neural correlates underlying policy matrix updates: in humans the changes of functional connectivity between DMN regions can be investigated following a temporal discounting experiment and in monkeys or rodents anterograde tracing can be used to study how homolog regions of the DMN present increased synaptic changes compare to other parts of the brain.
